---
title: 'IST 777: final project'
output:
  word_document: default
#  pdf_document: default
subtitle: Jeff Levesque
always_allow_html: yes
---

# Introductory / Descriptive Reports

```{r setup, include=FALSE}
## local variables
cwd = getwd()

## utility functions
if (!require('devtools')) install.packages('devtools', repos='http://cloud.r-project.org')
library('devtools')
devtools::install_local(paste0(cwd, '/packages/customUtility'))
library('customUtility')

## load packages
load_package(c('knitr', 'rmarkdown', 'BEST'))

## load dataset
allSchoolsReportStatus = read.csv(paste0(cwd, '/data/allSchoolsReportStatus.csv'))
reportSample = read.csv(paste0(cwd, '/data/reportSample.csv'))
usVaccines = read.csv(paste0(cwd, '/data/usVaccines.csv'))

## set seed
set.seed(777)
```

## Proportion of overall schools reported vaccination data

```{r}
library('ggplot2')
ggplot(
  allSchoolsReportStatus,
  aes(x=reported)
) +
geom_histogram(stat='count')
```
## Proportion of public and private schools reported vaccination data

```{r}
library('ggplot2')
ggplot(
  allSchoolsReportStatus,
  aes(x=reported)
) +
facet_grid(~pubpriv) +
geom_histogram(stat='count')
```

A significant proportion of public schools have reported more vaccinations than private school. Moreover, both schools who have not reported vaccinations, represent a very small marginal total:

```{r}
contingency = table(
  allSchoolsReportStatus$reported,
  allSchoolsReportStatus$pubpriv
)

rowSums(contingency)
colSums(contingency)

contingency
prop.table(contingency)
```

## U.S. vaccinations rates over time

Most vaccinations have stabilized. However, it appears that `HepB_BD` vacinations started much recently, and may require additional time to stabilize:

```{r}
library('reshape2')
df.m = melt(usVaccines,id='X')
ggplot(
  data = df.m, aes(x=X,y=value,colour=variable,group=variable)
) +
  geom_line()
```

In general the vacination rates reported by the World Health Organization for the United States, have a tendency to fluctuate initially. After some time, the rates seem to stabilize between 85 and 100. Furthermore, the `HepB_BD` rates does not show similar values, nor stability. Rather, the vacination endured a large growth period, then a decline. It is unknown whether the rates have stabilized, or future volatility could occur.

## Notable patterns in U.S. vaccinations rates over time

When using time series decomposition, with a `frequency=4`, there appears to be seasonal patterns for all vacination types. However, the trends are not consistent nor similar between the different vacination types.

```{r}
##
## custom decompose: allows custom title
##
## Note: https://stackoverflow.com/a/43054751
##
plot.decomposed.ts = function(x, title='') {
  xx = x$x
  if (is.null(xx))
    xx = with(x, if (type == 'additive')
      random + trend + seasonal
      else random * trend * seasonal)
  plot(
    cbind(
      observed = xx,
      trend = x$trend,
      seasonal = x$seasonal,
      random = x$random
    ),
    main=title
  )
}

data = subset(usVaccines, select = -c(X))
for(col in names(data)) {
  data.ts = ts(usVaccines[[col]], frequency = 4)
  data.decompose = decompose(data.ts, 'additive')
  plot.decomposed.ts(data.decompose, col)
}
```

# Public vs. Private School Comparisons

## Difference in overall proportions between public and private schools

### Frequentist t-test

The frequentist null hypothesis states there are no difference in proportions between the public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, the corresponding critical value 2.2e-16 < 0.05 = alpha value. This signifies a significant difference in proportions between public and private school.

```{r}
t.test(
  x=allSchoolsReportStatus[allSchoolsReportStatus$pubpriv=='PUBLIC', 1],
  y=allSchoolsReportStatus[allSchoolsReportStatus$pubpriv=='PRIVATE', 1]
)
```

### Bayes Factor (with HDI)

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. Specifically, the interval spans (-8.99e-06, 9.18e-06). Since this interval overlaps zero, there no credible evidence suggesting a mean difference between the report status of public and private school.

```{r}
library('BEST')
library('BayesFactor')

schoolReportOut = BESTmcmc(
  allSchoolsReportStatus['reported'][allSchoolsReportStatus$pubpriv=='PUBLIC', 1],
  allSchoolsReportStatus['reported'][allSchoolsReportStatus$pubpriv=='PRIVATE', 1],
  numSavedSteps=1000
)
plot(schoolReportOut)
```

## Overall vaccination rates between public and private schools

### Frequentist: t-test

The frequentist null hypothesis states there are no difference in proportions between the overall vaccination rates between public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, a critical value 0.085 > 0.05 = alpha value. Therefore, there is insufficient evidence to reject the null hypothesis.

```{r}
t.test(
  x=reportSample$allvaccs[reportSample$pubpriv == 'PUBLIC'],
  y=reportSample$allvaccs[reportSample$pubpriv == 'PRIVATE']
)
```

### Bayes Factor

The bayes factor provides an odds ratio in favor of the alternative hypothesis, namely there is a difference in proportions between the overall vaccination rates. In the below case, the odds ratio is 0.813:1 in favor of the alternative hypothesis. This suggests an inverted 1/0.813 => 1.23:1 ratio in favor of the null hypothesis. Though this ratio is greater than 1, it is significantly less than 3, and not worth mentioning.

```{r}
allVacsOut = anovaBF(
  formula = allvaccs ~ pubpriv,
  data=reportSample
)
summary(allVacsOut)
mcmcOut = posterior(allVacsOut, iterations=1000)
summary(mcmcOut)
```

Furthermore, the corresponding traceplot visually indicates some spikes at roughly 375, and almost 900. These outliers suggest that the markov chain monte carlo (MCMC), may not have converged. Thus using higher `iterations` could produce better results. However, the associated density curve, indicates a roughly normal distribution with no outliers.

```{r}
plot(mcmcOut[,'mu'])
```

### HDI

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. Specifically, the interval spans (0, 200). Since this interval includes zero, there no credible evidence suggesting a mean difference between the report status of public and private school.

```{r}
hist(mcmcOut)
abline(v=quantile(mcmcOut, c(0.025)), col='green')
abline(v=quantile(mcmcOut, c(0.975)), col='green')
```

## Medical exemptions between public and private schools

### Frequentist: t-test

The frequentist null hypothesis states there are no difference in proportions of medical exemptions between public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, a critical value 0.2774 > 0.05 = alpha value. Therefore, there is insufficient evidence to reject the null hypothesis.

```{r}
t.test(
  x=reportSample$medical[reportSample$pubpriv == 'PUBLIC'],
  y=reportSample$medical[reportSample$pubpriv == 'PRIVATE']
)
```

### Bayes Factor

The bayes factor provides an odds ratio in favor of the alternative hypothesis, namely there is a difference in proportions of medical exemptions between public and private schools. In the below case, the odds ratio is 0.179:1 in favor of the alternative hypothesis. This suggests an inverted 1/0.179 => 5.5:1 ratio in favor of the null hypothesis.

```{r}
medicalExOut = anovaBF(
  formula = medical ~ pubpriv,
  data=reportSample
)
summary(medicalExOut)
mcmcOut = posterior(medicalExOut, iterations=1000)
summary(mcmcOut)
```

Furthermore, the corresponding traceplot visually indicates some spikes at roughly 400, 700, and almost 900. These outliers suggest that the markov chain monte carlo (MCMC), may not have converged. Thus using higher `iterations` could produce better results. However, the associated density curve indicates a mostly normal distribution, while the central mode area curving slightly.

```{r}
plot(mcmcOut[,'mu'])
```

### HDI

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. In the below diagram, the interval includes zero, suggesting no credible evidence for a mean difference of medical exemptions between public and private schools.

```{r}
hist(mcmcOut)
abline(v=quantile(mcmcOut, c(0.025)), col='green')
abline(v=quantile(mcmcOut, c(0.975)), col='green')
```

## Religious/belief exemptions between public and private schools

### Frequentist: t-test

The frequentist null hypothesis states there are no difference in proportions of religious/belief exemptions between public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, a critical value 0.125 > 0.05 = alpha value. Therefore, there is insufficient evidence to reject the null hypothesis.

```{r}
t.test(
  x=reportSample$religious[reportSample$pubpriv == 'PUBLIC'],
  y=reportSample$religious[reportSample$pubpriv == 'PRIVATE']
)
```

### Bayes Factor

The bayes factor provides an odds ratio in favor of the alternative hypothesis, namely there is a difference in proportions of religious/belief exemptions between public and private schools. In the below case, the odds ratio is 0.442:1 in favor of the alternative hypothesis. This suggests an inverted 1/0.442 => 2.262:1 ratio in favor of the null hypothesis. Though this ratio is greater than 1, it is less than 3, and not worth mentioning.

```{r}
religiousOut = anovaBF(
  formula = religious ~ pubpriv,
  data=reportSample
)
summary(religiousOut)
mcmcOut = posterior(religiousOut, iterations=1000)
summary(mcmcOut)
```

Furthermore, the corresponding traceplot visually indicates some spikes at roughly 200, 250, 600, 900. These outliers suggest that the markov chain monte carlo (MCMC), may not have converged. Thus using higher `iterations` could produce better results. Furthermore, the associated density curve indicates a mostly normal distribution with a slighlty distorted peak.

```{r}
plot(mcmcOut[,'mu'])
```

### HDI

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. In the below diagram, the interval includes zero, suggesting no credible evidence for a mean difference of religious/belief exemptions between public and private schools.

```{r}
hist(mcmcOut)
abline(v=quantile(mcmcOut, c(0.025)), col='green')
abline(v=quantile(mcmcOut, c(0.975)), col='green')
```

## Predictive Analyses

### Predict public or private

Predicting a bivariate condition is possible using the logistic regression. By ensuring the dependent variable is a `factor`, and the `glm` function can be implemented:

```{r}
reportAdjustedSample = reportSample
reportAdjustedSample$pubpriv = as.numeric(reportAdjustedSample$pubpriv) - 1
pubpriv = glm(
  pubpriv ~ conditional + medical + religious,
  data = reportAdjustedSample,
  family = binomial()
)
summary(pubpriv)
```

The results logistic model have the following log odds results:

* for every unit change in conditional, the log odds of predicting a public school reporting vaccination changes by `-0.009361`.
* for every unit change in medical, the log odds of predicting a public school reporting vaccination changes by `-0.035439`.
* for every unit change in religious, the log odds of predicting a public school reporting vaccination changes by `0.0993`.

Moreover, each independent variable is found not significant. The religious independent variable had the lowest with `0.0993` > `0.05` = alpha level.

Additionally, the residuals vs. fitted plot indicate the the public school being overpredicted. This signifies highly positive residuals (as indicated above `666.6`), and generally a poor logistic regression. The corresponding Normal Q-Q plot does not provide much information other than ressembling the logistic pattern.

```{r}
plot(pubpriv)
```

### Predict conditional percentage

The following linear regression indicates that each independent variable, except `hepMiss` is statistically significant at an alpha = 0.05. Moreover, the general p-value = `2.2e-16` suggests that the overall model is statistically significant:

```{r}
conditional = lm(
  conditional ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  data = reportSample
)
summary(conditional)
```

When running the equivalent bayes factor, the posterior HDI indicates only `hepMiss` (-0.4065, 0.00463) overlaps zero. This indicates that the B-weights credibly differ from zero:

```{r}
bf.conditional = lmBF(
  data = reportSample,
  conditional ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  posterior = TRUE,
  iterations = 1000
)
summary(bf.conditional)
```

Furthemore, the bayes r-squared `0.718931` is less than `0.7227` from the conventional analysis:

```{r}
rsquared = 1 - (bf.conditional[, 'sig2'] / var(reportSample$conditional))
mean(rsquared)
quantile(rsquared, c(0.025))
quantile(rsquared, c(0.975))
```

Both approaches indicate a fairly strong relationship between the dependent variable (conditional), with the associated independent variables.

In an attempt to investigate the above coefficients, a chi-squared test was implemented as an omnibus test. The corresponding chi-squared test shows that `polMiss`, and `mmrMiss` are not statistically significant:

```{r}
anova(conditional, test='Chisq')
```

### Predict medical percentage

The following linear regression indicates that each independent variable, except `varMiss` is not statistically significant at an alpha = 0.05. Moreover, the general p-value = `1.995e-10` suggests that the overall model is statistically significant:

```{r}
conditional = lm(
  medical ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  data = reportSample
)
summary(conditional)
```

When running the equivalent bayes factor, the posterior HDI indicates only `hepMiss` (0.01741, 0.14185) does not overlap zero. This indicates a majority of variables have B-weights that do no credibly differ from zero:

```{r}
bf.conditional = lmBF(
  data = reportSample,
  medical ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  posterior = TRUE,
  iterations = 1000
)
summary(bf.conditional)
```

Furthemore, the bayes r-squared `0.07022636` is smaller than `0.07505` from the conventional analysis:

```{r}
rsquared = 1 - (bf.conditional[, 'sig2'] / var(reportSample$medical))
mean(rsquared)
quantile(rsquared, c(0.025))
quantile(rsquared, c(0.975))
```

However, both approaches indicate that the model does not predict a relationship between the dependent variable (medical), with the associated independent variables.

In an attempt to investigate the above coefficients, a chi-squared test was implemented as an omnibus test. The corresponding chi-squared test shows that `polMiss`, and `mmrMiss` are not statistically significant:

```{r}
anova(conditional, test='Chisq')
```

### Predict religious percentage

The following linear regression indicates `polMiss`, `mmrMiss`, and the corresponding intercept is not statistically significant at an alpha = 0.05. Moreover, the general p-value = `2.2e-16` suggests that the overall model is statistically significant:

```{r}
conditional = lm(
  religious ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  data = reportSample
)
summary(conditional)
```

When running the equivalent bayes factor, the posterior HDI indicates only `polMiss` (-0.03557, 0.2612), `mmrMiss` (-0.11431, 0.0825) does overlap zero.  This indicates that the B-weights credibly differ from zero:

```{r}
bf.conditional = lmBF(
  data = reportSample,
  religious ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  posterior = TRUE,
  iterations = 1000
)
summary(bf.conditional)
```

Furthemore, the bayes r-squared `0.7571835` is smaller than `0.7599` from the conventional analysis:

```{r}
rsquared = 1 - (bf.conditional[, 'sig2'] / var(reportSample$religious))
mean(rsquared)
quantile(rsquared, c(0.025))
quantile(rsquared, c(0.975))
```

Both approaches indicate a fairly strong relationship between the dependent variable (religious), with the associated independent variables.

In an attempt to investigate the above coefficients, a chi-squared test was implemented as an omnibus test. The corresponding chi-squared test shows that all variables were statistically significant:

```{r}
anova(conditional, test='Chisq')
```

## Big picture

Based on the earlier analysis, the logistic regression for predicting the `pubpriv` case on condition, was not a good model since none of the independent variables were significant. Furthemore, the corresponding diagnostic tools, such as the Residuals vs. Predicted indicated that the logistic model overpredicted.

Moreover, the percentage of students requiring a follow-up on vaccination records (i.e. conditional), were statistically significant with all factors, except `hepMiss`. Hepatitis vaccination is generally recommended by the CDC to infants between 6-23 months of age. Once a child has been provided treatment, this vaccine generally lasts a minimum of 10 years (hepA), and 20 years (hepB). This suggests if an infant receives proper vaccination shortly after birth, hepatitis should not be a problem through elementary school years.

Students with a religious/belief exemptions, were statistically significant with all factors, except `polMiss` and `mmrMiss`. Though early ideologies, including protest of the polio vaccine being created from animals may have existed, consistent state regulations have likely made this a nonsignificant issue today. Specifically, polio vaccinations is required by all 50 states and DC for children entering childcare and public school. Furthermore, religious/belief exemptions allowed students to avoid Measles/Mumps/Rubella vaccinations. However, in recent years, an alarming proportion of outbreaks have encouraged the removal of the exemption. State legislatures in New Jersey, New York, Iowa, Maine, and Vermont have proposed the removal of this exemption. Since polio vaccination is required by all states, this is equally non-significant as being able to avoid mmr vaccination altogether.

Finally, the percentage of student with a medical exemption was statistically significant with all factors except `varMiss`. Similar to the polio vaccination,  Varicella (i.e. chicken pox) vaccination is required by all 50 states including DC for all children entering childcare or public schools.

In general, the earlier analysis is a depiction on current societal beliefs, as well as a mirror representation of state laws and guidelines. It will be interesting in recent light, whether state legislatures in New Jersey, New York, Iowa, Maine, and Vermont, will be able to set a precedence among other states. Since ideologies are often re-evaluated - one may ask, should religious belief exempt an infant or child from receiving the best possible health option? Furthemore, religious convictions of one individual or family can jeopardize the health of another. Thus, such discussions and analysis is a fundamental journey to improving the overall health of a society.

## Sources

* https://freakonometrics.hypotheses.org/8210
* https://stats.idre.ucla.edu/r/dae/logit-regression/
* https://statistics.laerd.com/statistical-guides/one-way-anova-statistical-guide.php
* https://www.webmd.com/a-to-z-guides/prevention-15/vaccines/need-hepatitis-vaccines
* https://www.cdc.gov/hepatitis/hav/havfaq.htm
* https://www.cdc.gov/features/hepatitis-b-vaccine/index.html
* https://www.washingtonpost.com/religion/2019/02/21/some-anti-vaccination-parents-cite-religious-exemptions-measles-outbreaks-could-change-that/?utm_term=.d5c3ebef9153
* https://www.cdc.gov/vaccines/vpd/polio/public/index.html
* https://wwwnc.cdc.gov/eid/article/15/6/09-0087_article
* https://www.cdc.gov/vaccines/vpd-vac/varicella/basic-school-daycare-req.htm