---
title: 'IST 777: final project'
output:
  word_document: default
#  pdf_document: default
subtitle: Jeff Levesque
always_allow_html: yes
---

```{r setup, include=FALSE}
## local variables
cwd = getwd()

## utility functions
if (!require('devtools')) install.packages('devtools', repos='http://cloud.r-project.org')
library('devtools')
devtools::install_local(paste0(cwd, '/packages/customUtility'))
library('customUtility')

## load packages
load_package(c('knitr', 'rmarkdown'))

## load dataset
allSchoolsReportStatus = read.csv(paste0(cwd, '/data/allSchoolsReportStatus.csv'))
reportSample = read.csv(paste0(cwd, '/data/reportSample.csv'))
usVaccines = read.csv(paste0(cwd, '/data/usVaccines.csv'))
```

Proportion of public schools reported vaccination data:

```{r}
library('ggplot2')
ggplot(
  allSchoolsReportStatus,
  aes(x=reported)
) +
geom_histogram(stat='count')
```

Most vaccinations have stabilized. However, it appears that `HepB_BD` vacinations started much recently, and may require additional time to stabilize:

```{r}
library('reshape2')
df.m = melt(usVaccines,id='X')
ggplot(
  data = df.m, aes(x=X,y=value,colour=variable,group=variable)
) +
  geom_line()
```

In general the vacination rates reported by the World Health Organization for the United States, have a tendency to fluctuate initially. After some time, the rates seem to stabilize between 85 and 100. Furthermore, the `HepB_BD` rates does not show similar values, nor stability. Rather, the vacaination endured a large growth period, then a decline. It is unknown whether the rates have stabilized, or future volatility could occur.
