---
title: 'IST 777: final project'
output:
  word_document: default
#  pdf_document: default
subtitle: Jeff Levesque
always_allow_html: yes
---

# Introductory / Descriptive Reports

```{r setup, include=FALSE}
## local variables
cwd = getwd()

## utility functions
if (!require('devtools')) install.packages('devtools', repos='http://cloud.r-project.org')
library('devtools')
devtools::install_local(paste0(cwd, '/packages/customUtility'))
library('customUtility')

## load packages
load_package(c('knitr', 'rmarkdown', 'BEST'))

## load dataset
allSchoolsReportStatus = read.csv(paste0(cwd, '/data/allSchoolsReportStatus.csv'))
reportSample = read.csv(paste0(cwd, '/data/reportSample.csv'))
usVaccines = read.csv(paste0(cwd, '/data/usVaccines.csv'))

## set seed
set.seed(777)
```

## Proportion of overall schools reported vaccination data

```{r}
library('ggplot2')
ggplot(
  allSchoolsReportStatus,
  aes(x=reported)
) +
geom_histogram(stat='count')
```
## Proportion of public and private schools reported vaccination data

```{r}
library('ggplot2')
ggplot(
  allSchoolsReportStatus,
  aes(x=reported)
) +
facet_grid(~pubpriv) +
geom_histogram(stat='count')
```

A significant proportion of public schools have reported more vaccinations than private school. Moreover, both schools who have not reported vaccinations, represent a very small marginal total:

```{r}
contingency = table(
  allSchoolsReportStatus$reported,
  allSchoolsReportStatus$pubpriv
)

rowSums(contingency)
colSums(contingency)

contingency
prop.table(contingency)
```

## U.S. vaccinations rates over time

Most vaccinations have stabilized. However, it appears that `HepB_BD` vacinations started much recently, and may require additional time to stabilize:

```{r}
library('reshape2')
df.m = melt(usVaccines,id='X')
ggplot(
  data = df.m, aes(x=X,y=value,colour=variable,group=variable)
) +
  geom_line()
```

In general the vacination rates reported by the World Health Organization for the United States, have a tendency to fluctuate initially. After some time, the rates seem to stabilize between 85 and 100. Furthermore, the `HepB_BD` rates does not show similar values, nor stability. Rather, the vacination endured a large growth period, then a decline. It is unknown whether the rates have stabilized, or future volatility could occur.

## Notable patterns in U.S. vaccinations rates over time

When using time series decomposition, with a `frequency=4`, there appears to be seasonal patterns for all vacination types. However, the trends are not consistent nor similar between the different vacination types.

```{r}
##
## custom decompose: allows custom title
##
## Note: https://stackoverflow.com/a/43054751
##
plot.decomposed.ts = function(x, title='') {
  xx = x$x
  if (is.null(xx))
    xx = with(x, if (type == 'additive')
      random + trend + seasonal
      else random * trend * seasonal)
  plot(
    cbind(
      observed = xx,
      trend = x$trend,
      seasonal = x$seasonal,
      random = x$random
    ),
    main=title
  )
}

data = subset(usVaccines, select = -c(X))
for(col in names(data)) {
  data.ts = ts(usVaccines[[col]], frequency = 4)
  data.decompose = decompose(data.ts, 'additive')
  plot.decomposed.ts(data.decompose, col)
}
```

# Public vs. Private School Comparisons

## Difference in overall proportions between public and private schools

### Frequentist t-test

The frequentist null hypothesis states there are no difference in proportions between the public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, the corresponding critical value 2.2e-16 < 0.05 = alpha value. This signifies a significant difference in proportions between public and private school.

```{r}
t.test(
  x=allSchoolsReportStatus[allSchoolsReportStatus$pubpriv=='PUBLIC', 1],
  y=allSchoolsReportStatus[allSchoolsReportStatus$pubpriv=='PRIVATE', 1]
)
```

### Bayes Factor (with HDI)

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. Specifically, the interval spans (-8.99e-06, 9.18e-06). Since this interval overlaps zero, there no credible evidence suggesting a mean difference between the report status of public and private school.

```{r}
library('BEST')
library('BayesFactor')

schoolReportOut = BESTmcmc(
  allSchoolsReportStatus['reported'][allSchoolsReportStatus$pubpriv=='PUBLIC', 1],
  allSchoolsReportStatus['reported'][allSchoolsReportStatus$pubpriv=='PRIVATE', 1],
  numSavedSteps=1000
)
plot(schoolReportOut)
```

## Overall vaccination rates between public and private schools

### Frequentist: t-test

The frequentist null hypothesis states there are no difference in proportions between the overall vaccination rates between public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, a critical value 0.085 > 0.05 = alpha value. Therefore, there is insufficient evidence to reject the null hypothesis.

```{r}
t.test(
  x=reportSample$allvaccs[reportSample$pubpriv == 'PUBLIC'],
  y=reportSample$allvaccs[reportSample$pubpriv == 'PRIVATE']
)
```

### Bayes Factor

The bayes factor provides an odds ratio in favor of the alternative hypothesis, namely there is a difference in proportions between the overall vaccination rates. In the below case, the odds ratio is 0.813:1 in favor of the alternative hypothesis. This suggests an inverted 1/0.813 => 1:1.23 ratio in favor of the null hypothesis. Though this ratio is greater than 1, it is significantly less than 3, and not worth mentioning.

```{r}
allVacsOut = anovaBF(
  formula = allvaccs ~ pubpriv,
  data=reportSample
)
summary(allVacsOut)
mcmcOut = posterior(allVacsOut, iterations=1000)
summary(mcmcOut)
```

Furthermore, the corresponding traceplot visually indicates some spikes at roughly 375, and almost 900. These outliers suggest that the markov chain monte carlo (MCMC), may not have converged. Thus using higher `iterations` could produce better results. However, the associated density curve, indicates a roughly normal distribution with no outliers.

```{r}
plot(mcmcOut[,'mu'])
```

### HDI

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. Specifically, the interval spans (0, 200). Since this interval includes zero, there no credible evidence suggesting a mean difference between the report status of public and private school.

```{r}
hist(mcmcOut)
abline(v=quantile(mcmcOut, c(0.025)), col='green')
abline(v=quantile(mcmcOut, c(0.975)), col='green')
```

## Medical exemptions between public and private schools

### Frequentist: t-test

The frequentist null hypothesis states there are no difference in proportions of medical exemptions between public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, a critical value 0.2774 > 0.05 = alpha value. Therefore, there is insufficient evidence to reject the null hypothesis.

```{r}
t.test(
  x=reportSample$medical[reportSample$pubpriv == 'PUBLIC'],
  y=reportSample$medical[reportSample$pubpriv == 'PRIVATE']
)
```

### Bayes Factor

The bayes factor provides an odds ratio in favor of the alternative hypothesis, namely there is a difference in proportions of medical exemptions between public and private schools. In the below case, the odds ratio is 0.179:1 in favor of the alternative hypothesis. This suggests an inverted 1/0.179 => 5.5:1 ratio in favor of the null hypothesis.

```{r}
medicalExOut = anovaBF(
  formula = medical ~ pubpriv,
  data=reportSample
)
summary(medicalExOut)
mcmcOut = posterior(medicalExOut, iterations=1000)
summary(mcmcOut)
```

Furthermore, the corresponding traceplot visually indicates some spikes at roughly 400, 700, and almost 900. These outliers suggest that the markov chain monte carlo (MCMC), may not have converged. Thus using higher `iterations` could produce better results. However, the associated density curve indicates a mostly normal distribution, while the central mode area curving slightly.

```{r}
plot(mcmcOut[,'mu'])
```

### HDI

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. In the below diagram, the interval includes zero, suggesting no credible evidence for a mean difference of medical exemptions between public and private schools.

```{r}
hist(mcmcOut)
abline(v=quantile(mcmcOut, c(0.025)), col='green')
abline(v=quantile(mcmcOut, c(0.975)), col='green')
```

## Religious/belief exemptions between public and private schools

### Frequentist: t-test

The frequentist null hypothesis states there are no difference in proportions of religious/belief exemptions between public and private schools. The associated alternative hypothesis, suggests there exists a significant difference in proportions. Using the Welch two sample t-test, a critical value 0.125 > 0.05 = alpha value. Therefore, there is insufficient evidence to reject the null hypothesis.

```{r}
t.test(
  x=reportSample$religious[reportSample$pubpriv == 'PUBLIC'],
  y=reportSample$religious[reportSample$pubpriv == 'PRIVATE']
)
```

### Bayes Factor

The bayes factor provides an odds ratio in favor of the alternative hypothesis, namely there is a difference in proportions of religious/belief exemptions between public and private schools. In the below case, the odds ratio is 0.442:1 in favor of the alternative hypothesis. This suggests an inverted 1/0.442 => 2.262:1 ratio in favor of the null hypothesis. Though this ratio is greater than 1, it is less than 3, and not worth mentioning.

```{r}
religiousOut = anovaBF(
  formula = religious ~ pubpriv,
  data=reportSample
)
summary(religiousOut)
mcmcOut = posterior(religiousOut, iterations=1000)
summary(mcmcOut)
```

Furthermore, the corresponding traceplot visually indicates some spikes at roughly 200, 250, 600, 900. These outliers suggest that the markov chain monte carlo (MCMC), may not have converged. Thus using higher `iterations` could produce better results. Furthermore, the associated density curve indicates a mostly normal distribution with a slighlty distorted peak.

```{r}
plot(mcmcOut[,'mu'])
```

### HDI

The Bayesian HDI is a 95% density region, where points within this region are more credible than the remaining 5% distributed among the tails. In the below diagram, the interval includes zero, suggesting no credible evidence for a mean difference of religious/belief exemptions between public and private schools.

```{r}
hist(mcmcOut)
abline(v=quantile(mcmcOut, c(0.025)), col='green')
abline(v=quantile(mcmcOut, c(0.975)), col='green')
```

## Predictive Analyses

### Predict public or private

Predicting a bivariate condition is possible using the logistic regression. By ensuring the dependent variable is a `factor`, and the `glm` function can be implemented:

```{r}
reportAdjustedSample = reportSample
reportAdjustedSample$pubpriv = as.numeric(reportAdjustedSample$pubpriv) - 1
pubpriv = glm(
  pubpriv ~ conditional + medical + religious,
  data = reportAdjustedSample,
  family = binomial()
)
summary(pubpriv)
```

The results logistic model have the following log odds results:

* for every unit change in conditional, the log odds of predicting a public school reporting vaccination changes by `-0.009361`.
* for every unit change in medical, the log odds of predicting a public school reporting vaccination changes by `-0.035439`.
* for every unit change in religious, the log odds of predicting a public school reporting vaccination changes by `0.0993`.

Moreover, each independent variable is found not significant. The religious independent variable had the lowest with `0.0993` > `0.05` = alpha level.

Additionally, the residuals vs. fitted plot indicate the the public school being overpredicted. This signifies highly positive residuals (as indicated above `666.6`), and generally a poor logistic regression. The corresponding Normal Q-Q plot does not provide much information other than ressembling the logistic pattern.

```{r}
plot(pubpriv)
```

### Predict conditional percentage

The following linear regression indicates that each independent variable, except `hepMiss` is statistically significant at an alpha = 0.05. Moreover, the general p-value = `2.2e-16` suggests that the overall model is statistically significant:

```{r}
conditional = lm(
  conditional ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  data = reportSample
)
summary(conditional)
```

When running the equivalent bayes factor, the posterior HDI indicates only `hepMiss` (-0.4065, 0.00463) overlaps zero. This indicates that the B-weights credibly differ from zero:

```{r}
bf.conditional = lmBF(
  data = reportSample,
  conditional ~ dptMiss + polMiss + mmrMiss + hepMiss + varMiss,
  posterior = TRUE,
  iterations = 1000
)
summary(bf.conditional)
```

Furthemore, the bayes r-squared `0.718931` is less than `0.7227` from the conventional analysis:

```{r}
rsquared = 1 - (bf.conditional[, 'sig2'] / var(reportSample$conditional))
mean(rsquared)
quantile(rsquared, c(0.025))
quantile(rsquared, c(0.975))
```

In an attempt to investigate the above coefficients, a chi-squared test was implemented to determine whether one variable influences another. The corresponding chi-squared test shows that `polMiss`, and `mmrMiss` are not significant, and may be related:

```{r}
anova(conditional, test='Chisq')
```

## Sources

* https://freakonometrics.hypotheses.org/8210